---
title: "Querying Chicago Rideshare Data"
author: "Enoch Shin, Felix Stetsenko"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(RSocrata)
library(ggmap)
library(magrittr)
library(plotly)
library(feather)
library(sf)
knitr::opts_chunk$set(cache = TRUE,
                      echo=TRUE)
```


# Purpose

> This file ingests the transportation network provider (TNP) data from the City of Chicago data portal.


# Query City of Chicago Data

> City of Chicago data uses the Socrata API. You'll have to get an API key from the Chicago Data Portal.

> Navigate to the Socrata developer page (https://dev.socrata.com/foundry/data.cityofchicago.org/m6dm-c72p) for the Chicago data and scroll down. Click the button "sign up for an app token," and make an account to do so.

> Once you get your key, you can navigate to the appropriate data page on the City of Chicago Data Portal (https://data.cityofchicago.org/Transportation/Transportation-Network-Providers-Trips/m6dm-c72p). Click the "API" button in the top right corner and copy the `.json` link.

```{r, eval=FALSE}
url <- "https://data.cityofchicago.org/resource/m6dm-c72p.json"
mytoken <- "XXXXXXXXXXXXXXXX"
```

> We want to query Valentine's Day 2019 (February 14, 2019): the query is commented out below. Do some wrangling to make sure the desired variables are coded correctly.

```{r}
#######QUERY##############
# valday <- read.socrata(paste0(url, "?",
#"$where=trip_start_timestamp between '2019-02-14T00:00:00' and '2019-02-14T23:59:59'"),  
#app_token=mytoken)

# saveRDS(valday, "ChicagoCommute/RDA/valday.Rda")
valday <- readRDS("ChicagoCommute/RDA/valday.Rda")

# remove the coordinate variable, which is a list
valday <- valday %>% mutate_at(c("pickup_centroid_latitude",
                                 "pickup_centroid_longitude",
                                 "dropoff_centroid_longitude",
                                 "dropoff_centroid_latitude"), as.numeric) %>%
            dplyr::select(-contains(".coordinate"))
# write to feather for both the app folder and feather folder
write_feather(valday, "ChicagoCommute/feather/valday.feather")
write_feather(valday, "ChicagoCommute/app/feather/valday.feather")
```

# Univariate Glimpses into Data

```{r, warning=FALSE}
nrow(valday)

mosaic::favstats(valday$trip_miles)

mosaic::favstats(valday$trip_seconds)/60

ggplot(valday, aes(x=as.numeric(trip_miles))) + 
  geom_histogram(bins=35) +
  ggtitle("Number of miles traveled per trip") +
  xlab("Miles")

# count trips based on before or after sunset
# sunset on Valentine's Day 2019 was at 5:22 CST
## which is 11:22 PM UTC
sunset <- lubridate::as_datetime("2019-02-14 23:22:00")
sunset

day <- filter(valday, valday$trip_start_timestamp < sunset)
evening <- filter(valday, valday$trip_start_timestamp >= sunset)

nrow(day)/nrow(valday)
nrow(evening)/nrow(valday)
```


# Sample the data for smaller set

> The data in its raw form is too large to plot with leaflet. So, we'll take a random sample of n=500. 

```{r}
valday <- read_feather("ChicagoCommute/feather/valday.feather")
set.seed(32)
valday_small <- valday[sample(nrow(valday), 500),]

write_feather(valday_small, "ChicagoCommute/feather/valday_small.feather")
write_feather(valday_small, "ChicagoCommute/app/feather/valday_small.feather")
```


\newpage

# Appendix: experiments in visualization


## Making Simple Feature Object for Points

> Our main visualization uses polygons to draw lines between pickups and dropoffs. This is useful since we get an idea of where people are going for each trip. I initially tried to make simple feature objects instead of making polygons in leaflet, but this proved to be unnecessary and inferior since the `st_as_sf` function cannot handle NA values.


```{r, eval=FALSE}
valday <- read_feather("ChicagoCommute/feather/valday.feather")

pickups <- valday %>%
  drop_na("pickup_centroid_longitude") %>%
  drop_na("pickup_centroid_latitude")

dropoffs <- valday %>%
  drop_na("dropoff_centroid_longitude") %>%
  drop_na("dropoff_centroid_latitude")

pts_pickup <- st_as_sf(pickups, coords = c("pickup_centroid_longitude", 
                                           "pickup_centroid_latitude"), 
                       crs = 4326, na.fail = FALSE)

pts_dropoff <- st_as_sf(dropoffs, coords = c("dropoff_centroid_longitude",
                                             "dropoff_centroid_latitude"), 
                        crs = 4326, na.fail = FALSE)

# feather cannot write a list object, so use RDS
saveRDS(pts_pickup, "ChicagoCommute/RDA/pts_pickup.Rda")
saveRDS(pts_pickup, "ChicagoCommute/app/RDA/pts_pickup.Rda")

saveRDS(pts_dropoff, "ChicagoCommute/RDA/pts_dropoff.Rda")
saveRDS(pts_dropoff, "ChicagoCommute/app/RDA/pts_dropoff.Rda")
```

